Linear stability of magnetized massive protoplanetary disks

### 2014-04-05 paper/appendix.tex
tex updates
first draft pre spell check

\section{Relation between $Q$ and the Toomre parameter}\label{q3d2d}
The Toomre parameter defined for razor-thin disks is 
\begin{align}
  Q_\mathrm{2D}\equiv \frac{\kappa c_s}{\pi G\Sigma},  
\end{align}
where $\Sigma$ is the total column density. To relate our
self-gravity parameter $Q$ and $Q_\mathrm{2D}$, we replace $c_s$ by
$\overline{c_s}\equiv\int\rho c_s dz/\int\rho dz$, and replace
$\kappa$ by $\Omega$, so that 
\begin{align}
  Q_\mathrm{2D} = 2Qf \frac{\int_0^1
    \hat{\rho} \hat{c}_s
    d\hat{z}}{\left(\int_0^1 \hat{\rho} d\hat{z}\right)^2},   
\end{align}
where each term on the right-hand-side is non-dimensionalized (see
\S\ref{non-dim}). We display this relation in Fig. \ref{plot_q3d2d}
for the isothermal and polytropic disks.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/q2d_iso}
  \caption{Relation between the self-gravity parameter $Q$ and the
    classic Toomre parameter $Q_\mathrm{2D}$.   
    \label{plot_q3d2d}}
\end{figure}

\section{Reduction to linear hydrodynamics}\label{reduction}

Our task here is to remove the magnetic field and vertical velocity
perturbations from the linearized equations. Let us first define operators
\begin{align}
  D_0 = 1, \quad D_1 = \frac{\rho^\prime}{\rho} + \frac{d}{dz}, \quad
  D_2 = \frac{\rho^{\prime\prime}}{\rho} +
  \frac{2\rho^\prime}{\rho}\frac{d}{dz} + \frac{d^2}{d z^2},
\end{align}
and 
\begin{align}
  \overline{D}_0 = \eta D_0, \quad \overline{D}_1 = \eta^\prime D_0 + \eta
  D_1,\quad \overline{D}_2 = \eta^{\prime\prime} D_0 + 2\eta^\prime D_1 +
  \eta D_2. 
\end{align}
And we define the variables
\begin{align}
  &U \equiv \imgi\sigma\dvx - 2\Omega\dvy + \imgi k_x \w,\\
  &V \equiv \imgi\sigma\dvy + \frac{\kappa^2}{2\Omega}\dvx.
\end{align}

@@

Next, we examine separately the cases of a vertical field with
variable resistivity and that of a tilted field with uniform
resistivity. We remark that a similar procedure can be performed in
the general case of a tilted field with variable resistivity.

@@

\subsection{Tilted field with uniform resistivity}
Here we allow $\epsilon\neq 0$ but take $\eta$ to be constant. We
first obtain expressions for $\dbx$ and $\dby$. Differentiating
the $x$ momentum equation and replacing the resuting $\dbz^\prime$
using the divergence-free condition and $\dby^\prime$ using the $y$
momentum equation, we obtain an expression for $\dbx^{\prime\prime}$
which can be inserted into the $x$ induction equation. This gives
\begin{align}
  \imgi\sigma\dbx = B_z\dvx^\prime +
  \frac{\eta\mu_0\rho}{B_z}\left( D_1 U + \imgi\epsilon k_x D_0
  V\right).\label{bx_tilted}  
\end{align}
We can insert this into the $y$ induction equation to obtain
\begin{align}
  \imgi\sbar\dby = -B_y\Delta + B_z\dvy^\prime + \frac{\imgi
    S}{\sigma}\left[\frac{\eta\mu_0\rho}{B_z}\left( D_1 U + \imgi\epsilon k_x D_0
  V\right)\right] + \frac{\eta\mu_0\rho}{B_z}D_1V,\label{by_tilted}
\end{align} 
where we have also used the derivative of the $y$ momentum equation to
eliminate  $\dby^{\prime\prime}$. Recall $\Delta \equiv i k_x \dvx +
\dvz^\prime$, so that
\begin{align}
  \Delta = \imgi k_x \dvx + \frac{i}{\sigma}\left(\w^{\prime\prime} +
  \epsilon V^\prime\right) = -\left[\frac{\imgi\sigma W}{c_s^2} +
    \frac{\imgi\left(\ln{\rho}\right)^\prime}{\sigma}\left(\w^\prime +
    \epsilon V\right)\right], 
\end{align}
where the second equality results from the continuity equation.

@@

Similarly, we differentiate Eq. \ref{by_tilted} and use the $y$
momentum equation to eliminate $\dby^\prime$ to obtain
\begin{align}
0 = &v_A^2\dvy^{\prime\prime} + \frac{\imgi S}{\sigma}v_A^2
\dvx^{\prime\prime} +\frac{\imgi S}{\sigma}\eta D_2U +
\left\{\eta\left(D_2 - \frac{\epsilon k_x S}{\sigma} D_1\right) +
  \imgi\left[\frac{\epsilon^2v_A^2\left(\ln{\rho}\right)^{\prime\prime}}{\sigma}
    - \sbar\right]D_0\right\}V \notag\\ &+ \imgi\epsilon
  v_A^2\left\{\frac{\sigma}{c_s^2}\left[W^\prime -
    \left(\ln{c_s^2}\right)^\prime W\right] +
    \frac{1}{\sigma}\left[\left(\ln{\rho}\right)^\prime\w^{\prime\prime}
      +\left(\ln{\rho}\right)^{\prime\prime}\w^{\prime} \right]\right\}.    
\end{align}

@@

>>>>

\section{Relation between $Q$ and the Toomre parameter}\label{q3d2d}
The Toomre parameter defined for razor-thin disks is 
\begin{align} 
  Q_\mathrm{2D}\equiv \frac{\kappa c_s}{\pi G\Sigma},  
\end{align}
where $\Sigma$ is the total column density. To relate our
self-gravity parameter $Q$ and $Q_\mathrm{2D}$, we replace $c_s$ by
$\overline{c_s}\equiv\int\rho c_s dz/\int\rho dz$, and 
$\kappa$ by $\Omega$, giving
\begin{align}
  Q_\mathrm{2D} = 2Qf \frac{\int_0^1
    \hat{\rho} \hat{c}_s
    d\hat{z}}{\left(\int_0^1 \hat{\rho} d\hat{z}\right)^2},   
\end{align}
where each term on the right-hand-side is non-dimensionalized (see
\S\ref{non-dim}). Fig. \ref{plot_q3d2d} plots this relation 
for isothermal and polytropic disks.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/q2d_iso}
  \caption{Relation between the self-gravity parameter $Q$ used in
    this paper and the Toomre parameter $Q_\mathrm{2D}$ for razor-thin
    disks. 
    \label{plot_q3d2d}}
\end{figure}

\section{Reduction to linear hydrodynamics}\label{reduction} 
Our task here is to remove the magnetic field and vertical velocity
perturbations from the linearized equations. Let us first define operators
\begin{align}
  D_0 = 1, \quad D_1 = \frac{\rho^\prime}{\rho} + \frac{d}{dz}, \quad
  D_2 = \frac{\rho^{\prime\prime}}{\rho} +
  \frac{2\rho^\prime}{\rho}\frac{d}{dz} + \frac{d^2}{d z^2},
\end{align}
and 
\begin{align}
  \overline{D}_0 = \eta D_0, \quad \overline{D}_1 = \eta^\prime D_0 + \eta
  D_1,\quad \overline{D}_2 = \eta^{\prime\prime} D_0 + 2\eta^\prime D_1 +
  \eta D_2. 
\end{align}
And we define the variables
\begin{align}
  &U \equiv \imgi\sigma\dvx - 2\Omega\dvy + \imgi k_x \w,\\
  &V \equiv \imgi\sigma\dvy + \frac{\kappa^2}{2\Omega}\dvx.
\end{align}

@@

Next, we examine separately the cases of a vertical field with
variable resistivity and that of a tilted field with uniform
resistivity. (A similar procedure can be performed in
the general case of a tilted field with variable resistivity)

@@

\subsection{Tilted field with uniform resistivity}
Here we allow $\epsilon\neq 0$ but take $\eta$ to be constant. We
first obtain expressions for $\dbx$ and $\dby$. Differentiating
the $x$ momentum equation and replacing the resuting $\dbz^\prime$
using the divergence-free condition and $\dby^\prime$ using the $y$
momentum equation, we obtain an expression for $\dbx^{\prime\prime}$
which can be inserted into the $x$ induction equation. This gives
\begin{align}
  \imgi\sigma\dbx = B_z\dvx^\prime +
  \frac{\eta\mu_0\rho}{B_z}\left( D_1 U + \imgi\epsilon k_x D_0
  V\right).\label{bx_tilted}  
\end{align}
We can insert this into the $y$ induction equation to obtain
\begin{align}
  \imgi\sbar\dby = -B_y\Delta + B_z\dvy^\prime + \frac{\imgi
    S}{\sigma}\left[ B_z \dvx^\prime+ \frac{\eta\mu_0\rho}{B_z}\left( D_1 U + \imgi\epsilon k_x D_0
  V\right)\right] + \frac{\eta\mu_0\rho}{B_z}D_1V,\label{by_tilted}
\end{align} 
where we have also used the derivative of the $y$ momentum equation to
eliminate  $\dby^{\prime\prime}$. Recall $\Delta \equiv i k_x \dvx +
\dvz^\prime$, so that
\begin{align}
  \Delta = \imgi k_x \dvx + \frac{i}{\sigma}\left(\w^{\prime\prime} +
  \epsilon V^\prime\right) = -\left[\frac{\imgi\sigma W}{c_s^2} +
    \frac{\imgi\left(\ln{\rho}\right)^\prime}{\sigma}\left(\w^\prime +
    \epsilon V\right)\right], 
\end{align}
where the second equality results from the continuity equation.

@@

Similarly, we differentiate Eq. \ref{by_tilted} and use the $y$
momentum equation to eliminate $\dby^\prime$ to obtain
\begin{align}
0 = &v_A^2\dvy^{\prime\prime} + \frac{\imgi S}{\sigma}v_A^2
\dvx^{\prime\prime} +\frac{\imgi S}{\sigma}\eta D_2U +
\left\{\eta\left(D_2 - \frac{\epsilon k_x S}{\sigma} D_1\right) +
  \imgi\left[\frac{\epsilon^2v_A^2\left(\ln{\rho}\right)^{\prime\prime}}{\sigma}
    - \sbar\right]D_0\right\}V \notag\\ &+ \imgi\epsilon
  v_A^2\left\{\frac{\sigma}{c_s^2}\left[W^\prime -
    \left(\ln{c_s^2}\right)^\prime W\right] +
    \frac{1}{\sigma}\left[\left(\ln{\rho}\right)^\prime(\w^{\prime\prime}
      + \epsilon V^\prime)
      +\left(\ln{\rho}\right)^{\prime\prime}\w^{\prime} \right]\right\}.    
\end{align}

@@

%tilted field
% calculate b field vertical derivatives carefully
% try no velocity pert at bc
% try anti-aligned field

### 2014-04-05 paper/summary.tex
first draft pre spell check

For a vertical field, the requirement for the MRI is that   
its vertical wavelength $\lambda \lesssim H$. While $\lambda$ is independent of
the strength of self-gravity $Q$, the disk thickness $H=H(Q)$ 
decreases with increasing SG, which can reducie MRI growth rates. 
Thus, a sufficiently massive disk can potentially surpress the MRI. 
The MRI is also restricted to larger radial
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks. Finally, the condition $\lambda < H$ 
requries $f(Q)/\sqrt{\beta}$ to be sufficiently small, so weaker
fields are required to permit the MRI with increasing self-gravity.

Interestingly, for layered resistivity we do not find layered
perturbations when the disk is massive. This is 
consistent with MRI becoming vertically global with increasing self-gravity.    
For non-self-gravitating disks $\lambda\ll H$, so the MRI can be
restricted to regions of size $L<H$, i.e. an active layer. This is not
possible if $\lambda \sim H$, as found for massive disks. Hence we
find magnetic perturbations penetrate into the high-resistivity dead
zone (e.g. $Q=0.2$ in Fig. \ref{poly_layer}), and there is no distinct
boundary between active and dead layers. This suggests that the
picture of layered accretion  \citep[e.g.][]{fleming03} may not be applicable 
to self-gravitating disks.

%SG through 
We find MRI modes with radial scale of $\sim H$ can acquire  
density perturbations in massive but Toomre-stable disks. 
This occurs when the MRI is weak, for example with a strong field or
high resisitivty. We argue in that case $\lambda\sim H$, so the density
perturbation caused by the MRI can be enhanced by self-gravity.
%Our 
%interpretation involves the MRI to provide a seed density
%perturbation %in the absence of self-gravity.

@@

%further enhancing density perturbations. 
At this point it is worth mentioning previous non-linear simulations 
of magnetized self-gravitating galactic and circumstellar disks
\citep{kim03,fromang04a,fromang04}. These authors find self-gravity
does not enhance MRI density flucations significantly. However, they
employed ideal MHD simulations with gas-to-magnetic pressure ratios of
order $10^2$ to $10^3$. %MRI is basically incompressible 
This is qualitatively consistent with our results, as 
self-gravity is not expected to influence the MRI in this 
regime of $\beta$. On the other hand, \cite{fromang04} found
self-gravity makes MRI turbulence more coherent. This may be related
to our finding that small radial scale MRI is surpressed when
self-gravity is included in the background equilibria.

Physically, we expect MRI to interact with self-gravity when
their spatial scales are similar. Because self-gravity acts globally in the
vertical direction, for it to affect the MRI, future non-linear
simluations should consider parameter regimes in which the MRI is
vertically global. Indeed, in the setup of \cite{kim03}, the disk
scale height exceeds the MRI vertical wavelength and self-gravity has
little impact.

%GI/MRI: avoided crossings, relevance to simulations
Curiously, when GI and MRI are simultaneously supported, we find
unstable modes transition between MRI and GI. Such modes
have comparable potential and magnetic energy perturbations. These
transitions occur smoother with decreasing $\beta$
(Fig. \ref{compare_growth3}) or increasing $k_x$
(Fig. \ref{compare_growth3_Q01d2}). The latter implies that, in order
to capture magnetic-gravitational interactions represented by these
intermediate modes, non-linear simulations must resolve radial scales
smaller than the most unstable GI mode. For example, 
Fig. \ref{compare_growth3_Q01d2} suggest radial scales down to $\sim H/2$
should be resolved.

@@

We examined the effect of an additional azimuthal field, while
keeping the vertical field at fixed strength. In this case, we also
relaxed the equitorial symmetry condition applied previously and
considered the full disk column. Self-gravity affects the MRI
differently depending on the midplane symmetry of the MRI density
perturbation. Self-gravity destabilizes MRI modes where the magnetic
energy reaches a minimum at $z=0$, which have a symmetric
density perturbation in the limit $B_y\to0$. However, 
self-gravity stabilizes MRI modes where the magnetic energy maximizes
at $z=0$. These modes have an anti-symmetric density
perturbation in the limit $B_y\to0$. This stabilization effect is
stronger for increasing $B_y$. Previous linear calculations indicate
increased compressibility associated with a toroidal field stabilizes
the MRI \citep{kim03}. We conjecture that self-gravity can
further enhance this effect.   
%can't do WKB (can't tell symmetries)

@@

\emph{Beyond the shearing box.} The shearing box ignores the curvature
of toroidal field lines present in the global disk 
geometry. \cite{pessah05} have shown that new effects on the MRI arise
when the curvature of a super-thermal toroidal field is accounted for;
although \citeauthor{pessah05} focused on modes with large (small)
vertical (radial) wavenumbers, for which we do not expect self-gravity
to be important. Since compressibilty becomes important for strong
toroidal fields, the effect of self-gravity may become even more
significant when super-thermal toroidal fields are
considered. However, global disk models will be neccessary
to self-consistently probe this regime.

>>>>

For a vertical field, the requirement for the MRI is that   
its vertical wavelength $\lambda \lesssim H$. While $\lambda$ is independent of
the strength of self-gravity $Q$, the disk thickness $H=H(Q)$ 
decreases with increasing SG, which can reducie MRI growth rates. 
Thus, a sufficiently massive disk can potentially surpress the MRI. 
The MRI is also restricted to larger radial
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks. Finally, the condition $\lambda < H$ 
requries $f(Q)/\sqrt{\beta}$ to be sufficiently small, so weaker
fields are required to permit the MRI with increasing self-gravity.   
%since MRI is 3D, strong SG makes it 2D

Interestingly, for layered resistivity we do not find layered
magnetic perturbations when the disk is massive. This is 
consistent with the MRI becoming vertically global with increasing self-gravity.    
For non-self-gravitating disks $\lambda\ll H$, so the MRI can be
restricted to regions of size $L<H$, i.e. an active layer. This not
compatible with $\lambda \sim H$, as found for massive disks. Hence we
find magnetic perturbations penetrate into the high-resistivity dead
zone (e.g. $Q=0.2$ in Fig. \ref{poly_layer}), and there is no distinct
boundary between active and dead layers. This suggests that the
picture of layered accretion  \citep[e.g.][]{fleming03} may not be applicable 
to self-gravitating disks.

%SG through 
We find MRI modes with radial scales of $\sim H$ can acquire  
density perturbations in massive but Toomre-stable disks. 
This occurs when the MRI is weak, for example with a strong field or
high resisitivty. We argue in that case $\lambda\sim H$, so the MRI is
compressible and the associated density
perturbation can be enhanced by self-gravity.
%Our 
%interpretation involves the MRI to provide a seed density
%perturbation %in the absence of self-gravity.

@@

%further enhancing density perturbations. 
At this point it is worth mentioning previous non-linear simulations 
of magnetized self-gravitating galactic and circumstellar disks
\citep{kim03,fromang04a,fromang04}. These authors find self-gravity
did not enhance MRI density flucations significantly. However, they
employed ideal MHD simulations with gas-to-magnetic pressure ratios of
order $10^2$ to $10^3$. %MRI is basically incompressible 
This is qualitatively consistent with our results, as 
self-gravity is not expected to influence the MRI in this  
regime of $\beta$, except through the background state. 
%both `seed' and amplfication are small 
For example, \cite{fromang04} found
MRI turbulence is more coherent in self-gravitating disks. This may be related
to our finding that small radial scale MRI is surpressed when
self-gravity is included in the background equilibria.

Physically, we expect MRI to interact with self-gravity when
their spatial scales are similar. Because self-gravity acts globally in the
vertical direction, for it to affect the MRI, future non-linear
simluations should consider parameter regimes in which the MRI is
vertically global. %this makes MRI more compressible
Indeed, in the setup of \cite{kim03}, the disk
scale height exceeds the MRI vertical wavelength and self-gravity has
little impact.

%GI/MRI: avoided crossings, relevance to simulations
Curiously, when GI and MRI are simultaneously supported, we find
unstable modes transition between MRI and GI. There exists modes 
with comparable potential and magnetic energy perturbations, which 
demonstrates MRI and GI can interact. These 
transitions occur smoother with decreasing $\beta$
(Fig. \ref{compare_growth3}) or increasing $k_x$
(Fig. \ref{compare_growth3_Q01d2}). The latter implies that, in order
to capture the magneto-gravitational interactions represented by these
intermediate modes, non-linear simulations must resolve radial scales
smaller than the most unstable GI mode. For example, 
Fig. \ref{compare_growth3_Q01d2} suggest radial scales down to $\sim H/2$
should be well-resolved.

@@

We examined the effect of an additional azimuthal field, while
keeping the vertical field at fixed strength. In this case, we also
relaxed the equitorial symmetry condition applied previously and
considered the full disk column. Self-gravity affects the MRI
differently depending on its character. 
%midplane symmetry of the MRI 
%density
%perturbation. 
Self-gravity destabilizes MRI modes where the magnetic
energy has a minimum at $z=0$, these modes have a symmetric
density perturbation in the limit $B_y\to0$. However, 
self-gravity stabilizes MRI modes where the magnetic energy has a
maximum at $z=0$, these modes have an anti-symmetric density
perturbation in the limit $B_y\to0$. This stabilization effect is
stronger for increasing $B_y$. Previous linear calculations show that 
increased compressibility associated with a toroidal field stabilizes
the MRI \citep{kim03}. We conjecture that self-gravity 
further enhances this effect. Non-linear MRI simulations with strong
toroidal fields that neglect self-gravity may over-estimate the
strength of MRI turbulence.   
%can't do WKB (can't tell symmetries)

@@

\emph{Beyond the shearing box.} The shearing box ignores the curvature
of toroidal field lines present in the global disk 
geometry. \cite{pessah05} demonstrated new effects on the MRI 
when the curvature of a super-thermal toroidal field is accounted for;
although \citeauthor{pessah05} focused on modes with large (small)
vertical (radial) wavenumbers, for which we expect self-gravity
can be ignored. 
Since compressibilty becomes important for strong
toroidal fields, the effect of self-gravity on modes with $k_xH\sim1$
may become significant when super-thermal toroidal fields are
considered. However, global disk models will be neccessary
to self-consistently probe this regime.
